Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-23 numerical.tex
messed up notation change. need to fix by inspection.
restructured, moved early part of sec 5 to sec 4 as discussed. notation changes need further checking

We solve the linear problem by expanding the 
perturbations in Chebyshev polynomials $T_l$ up to $l=512$
and discretizing the equations on a grid with
$z\in[-\zmax,\zmax]$. Our standard boundary conditions at the vertical
boundaries is a free surface,  
\begin{align}
  \Delta P \equiv \delta P + \bm{\xi}\cdot\nabla P= 0 \quad \text{at } z=\pm\zmax,
\end{align}
% note that there are g_c terms in the boundary condition
where $\bm{\xi}$ is the Lagrangian displacement with meridional 
components $\xi_{x,z} = \ii\delta v_{x,z}/\sigma$. In some cases we
impose a rigid boundary at which $\delta v_z=0$.

@@

% Fig. \ref{iso_eigen_kx} compares the numerically obtained eigenvalues 
% for the fundamental mode and that calculated from 
% Eq. \ref{simple_growth}. We find good agreement in the
% wave frequency $\omega$ at all $\khat$ and in growth rates $\nu$ for
% $\khat\gtrsim 10$. Growth rates are over-estimated by our analysis but 
% are $O( h\Omega_k)$, which is consistent with the discussion in
% Appendix \ref{max_growth1}. The match in $\nu$ worsens for 
% decreasing $\khat$ due to the neglect of global radial structure in
% the analysis. Nevertheless, given the number of approximations made
% in our analysis, the agreement is satisfactory.

@@

The modes shown in Fig. \ref{compare_modes_iso_kx10} are `body modes'
and are inertial waves destabilized by vertical shear
\citepalias{barker15}. (The exception being the fastest growing
mode for $\zmax=7H$ which is a surface mode --- see the discussion
below.)  The `fundamental corrugation mode', marked by `f',
corresponds to the eigenvalue with the smallest $|\sigma|$. The
fundamental mode is robust  
against vertical boundary conditions as it appears at the same
position in Fig. \ref{compare_modes_iso_kx10} for all cases. Higher  
order modes, corresponding to increasing $|\omega|$, increasingly
deviates from the analytic line due to the imposed boundaries, 
unless $\zmax$ is increased.

>>>>

We solve the linear problem by expanding the 
perturbations in Chebyshev polynomials $T_l$ up to $l=512$
and discretizing the equations on a grid with
$z\in[-\zmax,\zmax]$. Our standard boundary conditions at the vertical
boundaries is a free surface,  
\begin{align}
  \Delta P \equiv \delta P + \bm{\xi}\cdot\nabla P= 0 \quad \text{at } z=\pm\zmax,
\end{align}
% note that there are g_c terms in the boundary condition
where $\bm{\xi}$ is the Lagrangian displacement with meridional 
components $\xi_{x,z} = \ii\delta v_{x,z}/\zeta$. In some cases we
impose a rigid boundary at which $\delta v_z=0$.

@@

% Fig. \ref{iso_eigen_kx} compares the numerically obtained eigenvalues 
% for the fundamental mode and that calculated from 
% Eq. \ref{simple_growth}. We find good agreement in the
% wave frequency $\omega$ at all $\khat$ and in growth rates $\sigma$ for
% $\khat\gtrsim 10$. Growth rates are over-estimated by our analysis but 
% are $O( h\Omega_K)$, which is consistent with the discussion in
% Appendix \ref{max_growth1}. The match in $\sigma$ worsens for 
% decreasing $\khat$ due to the neglect of global radial structure in
% the analysis. Nevertheless, given the number of approximations made
% in our analysis, the agreement is satisfactory.

@@

The modes shown in Fig. \ref{compare_modes_iso_kx10} are `body modes'
and are inertial waves destabilized by vertical shear
\citepalias{barker15}. (The exception being the fastest growing
mode for $\zmax=7H$ which is a surface mode --- see the discussion
below.)  The `fundamental corrugation mode', marked by `f',
corresponds to the eigenvalue with the smallest $|\zeta|$. The
fundamental mode is robust  
against vertical boundary conditions as it appears at the same
position in Fig. \ref{compare_modes_iso_kx10} for all cases. Higher  
order modes, corresponding to increasing $|\omega|$, increasingly
deviates from the analytic line due to the imposed boundaries, 
unless $\zmax$ is increased.

### 2015-04-23 setup.tex
restructured, moved early part of sec 5 to sec 4 as discussed. notation changes need further checking

\section{Governing equations and disk models}\label{setup}
We consider  an inviscid, non-self-gravitating three-dimensional (3D)
disk orbiting a central star of mass $M_*$.  Cylindrical
co-ordinates $(r,\phi, z)$ are centered on the star. The governing fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot{\left(\rho\bm{v}\right)} = 0,\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi_*,\\
  &\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  = - \gamma P
  \nabla\cdot\bm{v} - \Lambda % \frac{\rho}{t_c}\left(\frac{P}{\rho} -
    % \left.\frac{P}{\rho}\right|_{t=0}\right)
  , \label{full_energy}
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the velocity field (the
rotation frequency being $\Omega=v_\phi/r$), $P$
is the pressure, $\gamma$ is the constant adiabatic index, and $\Phi_*
= -GM_*(r^2 + z^2)^{-1/2}$ is the gravitational potential of the
central star and $G$ is the gravitational constant. 
In the energy equation (Eq.\ \ref{full_energy}) the sink term
$\Lambda$ includes non-adiabatic cooling and (negative) heating. 
The gas temperature $T$ obeys the ideal gas 
equation of state,  
\begin{align}
P = \frac{\mathcal{R}}{\mu}\rho T,
\end{align}
where $\mathcal{R}$ is the gas constant and $\mu$ is the mean molecular
weight.

@@

\section{Linear problem}\label{linear}

We consider axisymmetric perturbations to the above equilibria in the
form 
\begin{align}
  \rho \to \rho(r, z) + \delta\rho(r,z)\exp{\left( - \ii\zeta
      t\right)},    
\end{align}
and similarly for other fluid variables, where $\zeta = \omega + \ii
\sigma$ is a (generally) complex frequency, with $\omega$ and $\sigma$ being
the real frequency and growth rate, respectively.

% The linearized fluid equations in cylindrical co-ordinates are  
% \begin{align}
%   \ii\zeta\delta\rho &= \frac{\rho}{r}\frac{\p}{\p r}\left(r\delta v_r\right) + \frac{\p}{\p
%     z}\left(\rho\delta v_z\right)  + \hat{g}_c\frac{\p\rho}{\p r}\delta v_r,\label{full_mass1}\\
%   \ii\zeta\delta v_r &= \frac{1}{\rho}\frac{\p\delta P}{\p r} - 2\Omega
%   \delta v_\phi - \hat{g}_c\frac{\delta\rho}{\rho^2}\frac{\p P}{\p r},\\ 
%   \ii\zeta\delta v_\phi  &= \frac{\kappa^2}{2\Omega}\delta v_r +
%   r\frac{\p\Omega}{\p z}\delta v_z,\\
%   \ii \zeta\delta v_z &= \frac{1}{\rho}\frac{\p\delta P}{\p z} -
%   \frac{\delta \rho}{\rho^2}\frac{\p P}{\p z},\\
%   \ii\zeta\delta P &=\frac{\p P}{\p z}\delta v_z + \gamma P
%   \left[\frac{1}{r}\frac{\p}{\p r}\left(r\delta v_r\right) + \frac{\p
%       \delta v_z}{\p z}\right]\notag \\
%   &+ \frac{1}{t_c}\left(\delta P -  
%     \frac{P}{\rho}\delta\rho\right) + \hat{g}_c \frac{\p P}{\p r}\delta v_r,\label{full_energy1}
% \end{align}
% where the coefficients $\hat{g}_c=1$ are inserted to label terms
% associated with the background radial disk structure.

% This approach has been adopted
% in recent linear calculations of the VSI
% \cite{nelson13,mcnally14,barker15} because unstable modes have short
% radial lengthscales.

We consider perturbations with small radial length-scales 
so that we can ignore the radial variations in the background
structure and radial gradients. \citetalias{barker15} have validated this
approximation by comparison with the linear VSI in global cylindrical
disk geometry. We thus evaluate all coefficients in
Eq. \ref{full_mass1}---\ref{full_energy1} at the fiducial radius
$r=r_0$. However, the full vertical dependence of the coefficients are
retained. Then we can assume perturbations have spatial dependence in
the form 
\begin{align}
  \delta \rho(r,z) \to \delta\rho(z)\exp{\ii k_r r}.
\end{align}
We take $k_r>0$ without loss of generality, and further assume $k_rr\gg1$ so that curvature terms can also be 
ignored.

In order to emphasize the radially local approximation, we
introduce the local Cartesian co-ordinates $(x,y,z)$ corresponding to
the the local $(r,\phi,z)$ directions, respectively. Then writing
$\delta P /\rho \equiv W$ and $c_s^2\delta\rho/\rho\equiv Q$, the
linearized system of equations are 
\begin{align}
  \ii\zeta \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \hat{g}_c
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\zeta \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \zeta\delta v_y &= r\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\zeta\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \zeta W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x,\label{lin_energy}
  % -&\ii\zeta W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\zeta\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}
where we have made the notation change $k_r\to k_x$. 
It is understood that the quantities $\rho,\,c_s^2,\,\Omega,\kappa,\, 
\p_z\Omega,\, \p_r\ln{\rho} $ and $ \p_r P$ are evaluated at $r=r_0$
using their expressions in the global disk given in \S\ref{setup}. Thus, they are functions of $z$ only. 
Eq. \ref{lin_mass}---\ref{lin_energy} will form the basis of our numerical
study, but for analytical discussion
we will make further simplifications.

>>>>

\section{Governing equations and disk models}\label{setup}
We consider  an inviscid, non-self-gravitating three-dimensional (3D)
disk orbiting a central star of mass $M_*$.  Cylindrical
co-ordinates $(r,\phi, z)$ are centered on the star. The governing fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot{\left(\rho\bm{v}\right)} = 0,\label{full_mass}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi_*,\\
  &\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  = - \gamma P
  \nabla\cdot\bm{v} - \Lambda % \frac{\rho}{t_c}\left(\frac{P}{\rho} -
    % \left.\frac{P}{\rho}\right|_{t=0}\right)
  , \label{full_energy}
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the velocity field (the
rotation frequency being $\Omega=v_\phi/r$), $P$
is the pressure, $\gamma$ is the constant adiabatic index, and $\Phi_*
= -GM_*(r^2 + z^2)^{-1/2}$ is the gravitational potential of the
central star and $G$ is the gravitational constant. 
In the energy equation (Eq.\ \ref{full_energy}) the sink term
$\Lambda$ includes non-adiabatic cooling and (negative) heating. 
The gas temperature $T$ obeys the ideal gas 
equation of state,  
\begin{align}
P = \frac{\mathcal{R}}{\mu}\rho T,
\end{align}
where $\mathcal{R}$ is the gas constant and $\mu$ is the mean molecular
weight.

@@

\section{Linear problem}\label{linear} 
In this section we present the two sets of equations used to study the
axisymmetric linear stability of the above equilibria. We first give
the linearized equations in the radially-local approximation, which
are solved in our numerical study. From this set, we further derive a
simplified equation for vertically isothermal disks used for analytic
discussion in the next section.

\subsection{Radially local approximation}
We consider axisymmetric perturbations to the above equilibria in the
form 
\begin{align}
  \rho \to \rho(r, z) + \delta\rho(r,z)\exp{\left(\ii k_rr - \ii\zeta
      t\right)},    
\end{align}
and similarly for other fluid variables, where $\zeta = \omega + \ii
\sigma$ is a (generally) complex frequency, with $\omega$ and $\sigma$ being
the real frequency and growth rate, respectively, and $k_r$ is a real 
radial wavenumber. We take $k_r>0$ without loss of generality.

The VSI is expected to operate for perturbations with small radial
wavelengths \citepalias{nelson13,barker15}. We thus linearize 
Eq. \ref{full_mass}---\ref{full_energy} assuming perturbations 
in the above form with radially slowly-varying
amplitudes and $k_r r \gg 1$, so that the radial dependence of $\delta\rho(r,z)$, etc.,
may be neglected, as well as curvature terms. Henceforth we are
considering solutions at some fiducial radius $r=r_0$.  
%\citetalias{barker15} have validated this 
% approximation for the VSI by comparison with the linear problem in
% global cylindrical disk geometry.

In order to emphasize the radially local approximation, we introduce
the local Cartesian co-ordinates $(x,y,z)$ corresponding to the the
local $(r,\phi,z)$ directions, respectively.  
%We thus linearize 
%and replace  
% The linearized fluid equations in cylindrical co-ordinates are  
% \begin{align}
%   \ii\zeta\delta\rho &= \frac{\rho}{r}\frac{\p}{\p r}\left(r\delta v_r\right) + \frac{\p}{\p
%     z}\left(\rho\delta v_z\right)  + \hat{g}_c\frac{\p\rho}{\p r}\delta v_r,\label{full_mass1}\\
%   \ii\zeta\delta v_r &= \frac{1}{\rho}\frac{\p\delta P}{\p r} - 2\Omega
%   \delta v_\phi - \hat{g}_c\frac{\delta\rho}{\rho^2}\frac{\p P}{\p r},\\ 
%   \ii\zeta\delta v_\phi  &= \frac{\kappa^2}{2\Omega}\delta v_r +
%   r\frac{\p\Omega}{\p z}\delta v_z,\\
%   \ii \zeta\delta v_z &= \frac{1}{\rho}\frac{\p\delta P}{\p z} -
%   \frac{\delta \rho}{\rho^2}\frac{\p P}{\p z},\\
%   \ii\zeta\delta P &=\frac{\p P}{\p z}\delta v_z + \gamma P
%   \left[\frac{1}{r}\frac{\p}{\p r}\left(r\delta v_r\right) + \frac{\p
%       \delta v_z}{\p z}\right]\notag \\
%   &+ \frac{1}{t_c}\left(\delta P -  
%     \frac{P}{\rho}\delta\rho\right) + \hat{g}_c \frac{\p P}{\p r}\delta v_r,\label{full_energy1}
% \end{align}
% where the coefficients $\hat{g}_c=1$ are inserted to label terms
% associated with the background radial disk structure. 
% This approach has been adopted
% in recent linear calculations of the VSI
% \cite{nelson13,mcnally14,barker15} because unstable modes have short
% radial lengthscales. 
%We consider perturbations with small radial length-scales 
%so that we can ignore the radial variations in the background
%structure and radial gradients. 
%We thus evaluate all coefficients in
%Eq. \ref{full_mass1}---\ref{full_energy1} at the fiducial radius
%$r=r_0$. However, the full vertical dependence of the coefficients are
%retained. % Then we can assume perturbations have spatial dependence in
% the form 
% \begin{align}
%   \delta \rho(r,z) \to \delta\rho(z)\exp{\ii k_r r}.
% \end{align}
% We take $k_r>0$ without loss of generality, and further assume $k_rr\gg1$ so that curvature terms can also be 
% ignored.       
Then writing $\delta P /\rho \equiv W$ and $c_s^2\delta\rho/\rho\equiv Q$, the
linearized system of equations are 
\begin{align}
  \ii\zeta \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \hat{g}_c
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\zeta \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \zeta\delta v_y &= r\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\zeta\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \zeta W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x,\label{lin_energy}
  % -&\ii\zeta W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\zeta\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}
where we have made the notation change $k_r\to k_x$. 
It is understood that the quantities $\rho,\,c_s^2,\,\Omega,\kappa,\, 
\p_z\Omega,\, \p_r\ln{\rho} $ and $ \p_r P$ are evaluated at $r=r_0$
using their expressions in the global disk given in \S\ref{setup}. Thus, they are functions of $z$ only. 
Eq. \ref{lin_mass}---\ref{lin_energy} will form the basis of our numerical
study, but for analytical discussion
we will make further simplifications, as discussed next.
